---
title: "00 Setup — Spatial Equity of Social Infrastructure in Aarhus"
author: "Réka Forgó"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 3
---

```{r setup, include=FALSE}
library(knitr)
library(rmdformats)
opts_chunk$set(echo = TRUE, cache = FALSE, prompt = FALSE,
               tidy = TRUE, comment = NA, message = FALSE, warning = FALSE)
opts_knit$set(width = 75)
```

# Purpose

Environment setup: create the folder structure, load packages, and fix the CRS
conventions. Run once at the start of a session, then move to `01_get_data.Rmd`.

Reproduction note: to re-run the analysis without re-downloading anything, take
the processed data from the GitHub Release, unzip it into `data/processed/`,
and knit `02_analysis.Rmd`. Script `01` is only needed to rebuild that data
from source.

# 1. Project structure

```{r project-structure}
library(here)
library(fs)
library(purrr)

required_dirs <- c("data/raw", "data/processed", "R",
                   "output/figures", "output/maps", "report")
walk(required_dirs, ~ dir_create(here(.x)))
dir_tree(here(), recurse = 1)
```

# 2. Packages

```{r install-pacman, eval = FALSE}
install.packages("pacman")
```

```{r packages}
library(pacman)
p_load(sf, tidyverse, here, fs,
       osmdata, sfarrow,            # acquisition
       sfnetworks, tidygraph, units, # network distances
       spatstat, spdep,             # point patterns + autocorrelation
       tmap, mapview, leaflet, htmltools, htmlwidgets, # maps
       ggrepel, scales)             # plotting helpers
```

# 3. CRS conventions

| Purpose | EPSG | Name |
|---|---|---|
| OSM input / Leaflet output | 4326 | WGS 84 (lon/lat, degrees) |
| All analysis (buffers, area, distance) | 25832 | ETRS89 / UTM 32N (metres) |

All analysis runs in EPSG:25832; reproject to 4326 only for Leaflet.

```{r crs-constants}
crs_wgs <- 4326
crs_dk  <- 25832
```

# 4. Sanity check

```{r sanity}
aarhus_bbox <- st_bbox(c(xmin = 9.95, ymin = 56.05, xmax = 10.35, ymax = 56.27),
                       crs = st_crs(crs_wgs))
aarhus_bbox_sf <- st_as_sfc(aarhus_bbox) %>% st_transform(crs_dk)
mapview(aarhus_bbox_sf, alpha.regions = 0.1)
```

If the map lands on Aarhus, the environment is ready.

# Decisions documented here

- R version: `r R.version.string`
- Analysis CRS: EPSG:25832; I/O CRS: EPSG:4326
- Network-distance engine: `sfnetworks::st_network_cost()` on the OSM walking graph
- Euclidean-buffer engine: `sf::st_buffer()`
- Intermediates cached to `data/processed/`; scripts read from disk, so expensive
  steps run once and are skipped on reruns.
